---
title: "How Has My Name Changed in Popularity?"
author: ""
format:
  html:
    toc: true
    code-fold: true
server: shiny
execute:
  warning: false
  message: false
---

## Research question

How has the popularity of the selected name changed during the ten most recent years available in the U.S. Social Security Administration baby-name data?

```{r}
#| context: setup
#| echo: false

textInput(
  inputId = "name",
  label = "Enter a name:",
  value = "Olivia",
  width = "300px"
)

actionButton(
  inputId = "submit_name",
  label = "Update charts"
)
```

## Data

This analysis uses baby-name records from the [U.S. Social Security Administration](https://www.ssa.gov/oact/babynames/). Each row contains a name, sex, year, and number of babies given that name.

```{r}
#| context: setup
#| label: setup

library(tidyverse)
library(shiny)

zip_url <- "https://www.ssa.gov/oact/babynames/names.zip"
zip_file <- "names.zip"
data_dir <- "ssa-names"

if (!dir.exists(data_dir)) {
  dir.create(data_dir)
}

ssa_available <- if (file.exists(zip_file)) {
  TRUE
} else {
  tryCatch({
    download.file(zip_url, zip_file, mode = "wb")
    TRUE
  }, error = \(error) FALSE)
}

if (ssa_available) {
  if (length(list.files(data_dir, pattern = "^yob[0-9]{4}\\.txt$")) == 0) {
    unzip(zip_file, exdir = data_dir)
  }

  name_files <- list.files(
    data_dir,
    pattern = "^yob[0-9]{4}\\.txt$",
    full.names = TRUE
  )

  baby_names <- map_dfr(name_files, \(file) {
    read_csv(
      file,
      col_names = c("name", "sex", "births"),
      show_col_types = FALSE
    ) |>
      mutate(year = as.integer(str_extract(basename(file), "[0-9]{4}")))
  })

  source_note <- "the SSA data"
} else {
  # This older mirror keeps the document renderable if SSA blocks downloads.
  fallback_file <- "babynames.rda"
  if (!file.exists(fallback_file)) {
    download.file(
      "https://raw.githubusercontent.com/hadley/babynames/master/data/babynames.rda",
      fallback_file,
      mode = "wb"
    )
  }

  load(fallback_file)
  baby_names <- babynames |>
    transmute(year, name, sex, births = n)
  source_note <- "the packaged babynames mirror"
}
```

This document is using `r source_note`. The mirror is an older copy, so if the SSA download is available it will automatically be used instead.

## Prepare the data

The analysis uses the ten latest complete years in the dataset. Rank is calculated separately for each year and sex because the SSA data reports names by sex.

```{r}
#| context: server
latest_year <- max(baby_names$year)
first_year <- latest_year - 9

ranked_names <- baby_names |>
  filter(year >= first_year, year <= latest_year) |>
  group_by(year, sex) |>
  arrange(desc(births), name, .by_group = TRUE) |>
  mutate(
    rank = row_number(),
    popularity_share = births / sum(births)
  ) |>
  ungroup()

submitted_name <- eventReactive(
  input$submit_name,
  str_squish(input$name),
  ignoreNULL = FALSE
)

name_trend <- reactive({
  my_name <- submitted_name()

  trend <- ranked_names |>
    filter(str_to_lower(name) == str_to_lower(my_name)) |>
    select(year, name, sex, births, rank, popularity_share)

  trend |>
    complete(
      year = first_year:latest_year,
      sex,
      fill = list(births = 0, popularity_share = 0)
    ) |>
    mutate(name = my_name)
})

output$births_plot <- renderPlot({
  trend <- name_trend()
  validate(need(nrow(trend) > 0, "No records found for this name."))

  ggplot(trend, aes(x = year, y = births, color = sex)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_x_continuous(breaks = first_year:latest_year) +
    labs(
      title = paste("Births with the name", submitted_name()),
      x = "Year",
      y = "Number of babies",
      color = "Sex"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

output$rank_plot <- renderPlot({
  trend <- name_trend()
  validate(need(nrow(trend) > 0, "No records found for this name."))

  ggplot(trend, aes(x = year, y = rank, color = sex)) +
    geom_line(linewidth = 1, na.rm = TRUE) +
    geom_point(size = 2, na.rm = TRUE) +
    scale_x_continuous(breaks = first_year:latest_year) +
    scale_y_reverse() +
    labs(
      title = paste("Rank of the name", submitted_name()),
      x = "Year",
      y = "Rank (1 is most popular)",
      color = "Sex"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

output$summary_table <- renderTable({
  trend <- name_trend()
  validate(need(nrow(trend) > 0, "No records found for this name."))

  trend |>
    mutate(popularity_share = scales::percent(popularity_share, accuracy = 0.001)) |>
    arrange(sex, year)
})

output$conclusion_text <- renderText({
  trend <- name_trend() |> filter(births > 0)
  validate(need(nrow(trend) > 0, "No records found for this name."))

  summaries <- trend |>
    group_by(sex) |>
    group_split() |>
    map_chr(\(df) {
      sex_label <- if (unique(df$sex) == "F") "female" else "male"
      start <- df |> filter(year == min(year))
      end <- df |> filter(year == max(year))
      peak <- df |> slice_max(births, n = 1, with_ties = FALSE)
      direction <- if (end$births > start$births) "increased" else if (end$births < start$births) "decreased" else "remained steady"

      paste0(
        "For ", sex_label, " babies, the name ", submitted_name(), " ", direction, " from ",
        format(start$births, big.mark = ","), " births (rank #", start$rank, ") in ", start$year,
        " to ", format(end$births, big.mark = ","), " births (rank #", end$rank, ") in ", end$year, ". ",
        "Its peak popularity was in ", peak$year, " with ",
        format(peak$births, big.mark = ","), " births (rank #", peak$rank, ")."
      )
    })

  paste(summaries, collapse = "\n\n")
})
```

## Number of babies

```{r}
plotOutput("births_plot")
```

## Rank

A smaller rank means greater popularity. The reversed y-axis makes movement toward rank 1 appear higher on the graph.

```{r}
plotOutput("rank_plot")
```

## Summary table

```{r}
tableOutput("summary_table")
```

## Patterns to discuss

Use the graphs and table to answer these questions:

- Did the number of babies given this name increase, decrease, or stay fairly stable?
- Did the name's rank change in the same direction as its number of births?
- Does the name appear for more than one sex? If so, is one trend different from the other?
- Were there any sudden increases or decreases?
- Did the name become more or less common relative to all births?
- What cultural, historical, or generational explanation might help explain the pattern?

## Conclusion

```{r}
textOutput("conclusion_text")
```
